Deterministic Bayesian inference methods for the Naomi model

### 2023-04-14 src/dev_scale-grid/scale-grid.Rmd
Almost finished with slides for group meeting #25

C_plot <- C_df %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

@@

```{r}
start <- Sys.time()
quad <- fit_aghq(tmb_inputs, basegrid = base_grid)
end <- Sys.time()

@@

ggplot(tv_df, aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 20, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()
```

@@

Let's start by keeping `s = 5` dimensions.
We can create a dense GHQ grid with `k = 2` on 5 dimensions.

```{r}
s <- 5
d <- dim(E)[1]
gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = 2) 
```

This grid has $2^5 = 32$ nodes:

@@

C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C_s[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )

{C_plot + theme(legend.position = "bottom")} + C_s_plot
```

@@

Adaptive Gauss-Hermite quadrature involves

>>>>

C_plot <- C_df %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

@@

```{r}
start <- Sys.time()
quad <- fit_aghq(tmb_inputs, k = 1, basegrid = base_grid)
end <- Sys.time()

@@

(tv_plot <- ggplot(tv_df, aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 20, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal())

ggsave("tv-plot.png", plot = tv_plot, h = 4, w = 6.25)
```

@@

Let's start by keeping `s = 8` dimensions.
We can create a dense GHQ grid with `k = 3` on 8 dimensions.

```{r}
s <- 8
d <- dim(E)[1]
gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = 3) 
```

This grid has $3^8 = 6561$ nodes:

@@

C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
  ggplot(aes(x = as.factor(Var1), y = as.factor(Var2), fill = value)) +
  labs(x = "", y = "", fill = "C_s[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "none"
  )

(reduced_rank_plot <- {C_plot + theme(legend.position = "none") + labs(subtitle = "Full rank")} + {C_s_plot + labs(subtitle = "Reduced rank")})

ggsave("reduced-rank-plot.png", reduced_rank_plot, h = 4, w = 6.25)
```

@@

### 2023-04-14 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Almost finished with slides for group meeting #25

1. It runs quickly
2. It is flexible enough to be compatible with the model
3. We don't have better options

@@

Goal: approximate this integral for Naomi better than \texttt{TMB} does. In doing so, more accurately reflect uncertainty over hyperparameters

@@

* Pretend the posterior distribution is a Gaussian! Then it's easy

@@

1. Optimising a function to obtain `opt_bfgs`
2. Taking the mode `opt_bfgs$mode` and the Hessian at the mode `opt_bfgs$hessian`

@@

# Small interlude

* In spatio-temporal statistics we have data indexed by space and time
* To model this data, we use random effects also indexed by space and time
* There might be a lot of spatio-temporal locations

@@

@@

* Gauss-Hermite quadrature is a method for picking nodes and weights based on the theory of polynomial interpolation
* It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
* The "adaptive" part means the nodes and weights are changed depending on the integrand. This makes sense, especially when the integrand is also a function of the data
* Implemented by the `aghq` package [@stringer2021implementing]

@@

# The plan

* Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
* Use `TMB` for writing the model in C++ and implementing the Laplace approximation via automatic differentitation
* Use a adaptive Gauss-Hermite quadrature to integrate the hyperparameters

@@

* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
* Proposed solution: use principle components analysis

@@

* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work
  * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Let me know if you'd be up for being an early reader!
* Code for this project is at `athowes.github.io/elgm-inf`

>>>>

1. It runs quickly
2. It is flexible enough to be compatible with the model
3. We don't have better viable options

@@

Goal: approximate this integral for Naomi better than \texttt{TMB} does, without taking too long. In doing so, more accurately reflect uncertainty over hyperparameters

@@

* If you pretend the posterior distribution is a Gaussian, computation is easy

@@

1. Optimising a function
2. Taking the mode and the Hessian at the mode

@@

# Which parameters should we treat as Gaussian?

* In spatio-temporal statistics we have data indexed by space and time
* We use random effects also indexed by space and time to model this data
* Spatio-temporal fields can be big
* Willing to make assumptions about how things vary over spacetime

$\implies$ treat the latent field parameters as Gaussian! That's the majority of the integral done.

@@

* You evaluate the function at the nodes, then do a weighted sum to calculate your integral

@@

* Gauss-Hermite quadrature is a method for picking nodes and weights based on the theory of polynomial interpolation
* It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution
* The "adaptive" part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data
* Implemented by the `aghq` package [@stringer2021implementing]

@@

# The plan?

::: {.column width=.55}

* Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
* Use `TMB` for writing the model in C++ and implementing the Laplace approximation via automatic differentitation
* Use adaptive Gauss-Hermite quadrature to integrate the hyperparameters

::: {.column width=.45}

```{r, echo=FALSE, fig.cap="One of the best engineering schools in Canada", out.width = '90%'}
knitr::include_graphics("the-plan.png")
```

@@

* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
* Proposed solution use principal components analysis and keep only first $s < 24$ dimensions
* We will use $k = 3$ points per dimension

```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what this looks like for the 2D example, keeping only the first principal component, $s = 1$."}
gg3 <- gg
mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)

```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$.", out.width = '80%'}
knitr::include_graphics("depends/tv-plot.png")
```

```{r, echo=FALSE, fig.cap="With 8 dimenions, the covariance matrix is accurately reproduced.", out.width = '80%'}
knitr::include_graphics("depends/reduced-rank-plot.png")
```

# Comparison of posterior distributions

* Use test data from Malawi
* Run HMC for as long as it takes to get a good answer
* Compare output of `TMB` with `aghq`: which is more similar to HMC?
* There are many ways you might do this: I'll show you Kolmogorov-Smirnov test results

Worse case TMB plot demonstrating KS test

Overall KS results plot

# Conclusions

@@

* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work
  * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Code and notebooks for this project are at `athowes.github.io/elgm-inf`

Let me know if you'd be up for being an early reader!

# Risk group retrospective

```{r, echo=FALSE, fig.cap="Available at \\texttt{athowes.github.io}", out.width = '90%'}
knitr::include_graphics("blog-post.png")
```
